Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-06-30 introduction.tex
Tuned sentence about striking balance between latency and throughput
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1228 ccdd05d5-56c1-461a-bf98-85f6876523b4
Tuned sentence about sending prompt alerts to the astronomical community
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1229 ccdd05d5-56c1-461a-bf98-85f6876523b4

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW{} searches.  The most obvious example of the latter is
the time domain (\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.  \citet{shaunIIR} explored decomposing
\CBC\ matched filters into banks of infinite impulse response (\textsc{iir})
filters as a means to reduce the computational cost of latency-free data analysis.  They
demonstrate a factor of 50 improvement in floating point operations per second
over the standard \TD\ method.  We will significantly improve upon that number.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \citep{Cannon:2010p10398}. We will use both aspects to demonstrate
that a very low latency detection statistic is possible with current computing
resources.  Assuming the other technical sources of latency can be reduced
significantly, this should allow the possibility for prompt alerts to be sent
to the astronomical community.

>>>>

This work will focus on reducing the latency of trigger production.  Data analysis
strategies for advance detection of \CBC{}s will have to strike a balance between latency
and throughput. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW{} searches.  The most obvious example of the latter is
the time domain (\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.  \citet{shaunIIR} explored decomposing
\CBC\ matched filters into banks of infinite impulse response (\textsc{iir})
filters as a means to reduce the computational cost of latency-free data analysis.  They
demonstrate a factor of 50 improvement in floating point operations per second
over the standard \TD\ method.  We will significantly improve upon that number.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \citep{Cannon:2010p10398}. We will use both aspects to demonstrate
that a very low latency detection statistic is possible with current computing
resources.  Assuming the other technical sources of latency can be reduced
significantly, this could make it possible to send prompt alerts to the
astronomical community.
